---
title: "Data preparation"
date: "10 November 2020"
author: 
- "Rachel"
- "Eliane"
output: html_document
---

## Creek data 
```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load data
f1 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/18_4_data_Welk_etal_PlosOne.csv", sep=";", header=TRUE)
f2 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Plant species percent cover data.csv", sep="\t", header=TRUE)
f3 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Plant aboveground biomass carbon and nitrogen.csv", sep="\t", header=TRUE)
f4 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Plant aboveground biomass data.csv", sep="\t", header=TRUE)
f5 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Plant traits.csv", sep="\t", header=TRUE)
f6 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Percent light penetration.csv", sep="\t", header=TRUE)
f7 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Soil carbon.csv", sep="\t", header=TRUE)
f8 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Soil nitrate and ammonium.csv", sep="\t", header=TRUE)
f9 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek/e120_Soil nitrogen.csv", sep="\t", header=TRUE)

f_1 <- merge(f1, f2, by=c("Plot"), all=T)
f_2 <- merge(f_1, f3, by=c("Plot","Year"), all=T)


```


## Data from Welk et al. (2014) Biotic Interactions Overrule Plant Responses to Climate, Depending on the Species' Biogeography

We downloaded the data from https://idata.idiv.de/ddm/Data/ShowData/18



```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load data
raw_data <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/Welk/18_4_data_Welk_etal_PlosOne.csv", sep=";", header=TRUE)

```


## Data from Seabloom et al. (2020) 

# Grassland ecosystem recovery after soil disturbance depends on nutrient supply rate (Ecology Letters, 2020, 23: 1756–1765)

Recovery of a grassland for two decades following an intensive agricultural disturbance under ambient and elevated nutrient conditions:
- Productivity returned to pre-disturbance levels quickly under ambient nutrient conditions, but nutrient addition slowed this recovery
- Diversity did not change for 15 years, afterwards diversity began to increase under ambient conditions 


```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load data
raw_data <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek_Seabloom/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data.csv", sep=",", header=TRUE)

raw_data2 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek_Seabloom/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data-origin-group.csv", sep=",", header=TRUE) #"disk.trt","trt","origin"

library(tidyr)
raw_data2_2 <- unique(spread(raw_data2, "origin", "mass.above"))

test <- merge(raw_data, raw_data2_2, by=c("field"  ,    "exp"   ,     "plot"   ,    "year"   ,    "disk"     ,   "ntrt"     ,  "nadd"     ,  "other.add" ,"dur"), all=T)

raw_data3 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek_Seabloom/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data-life-group.csv", sep=",", header=TRUE) #"disk.trt", "trt", 

raw_data4 <- read.table("~/Desktop/SEM_UZH/Swiss_SEM/data_preparation/CedarCreek_Seabloom/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data-fun-group.csv", sep=",", header=TRUE)

test <- merge(raw_data, raw_data4, by=c("field"  ,    "exp"   ,     "plot"   ,    "year"   ,    "disk"     ,   "ntrt"     ,  "nadd"     ,  "other.add" ,"dur","mass.above"), all=T)




```



