---
author: "Frank Pennekamp"
title: "Introduction to structural equation modelling"
subtitle: "Day 2: Advanced modelling"
institute: 
- Department of Evolutionary Biology and Environmental Sciences, University of Zurich
date: "1/17/2024"
date-format: long
format:
  revealjs: 
    smaller: false
    scrollable: false
    theme: simple
    slide-number: true
    show-slide-number: all
    code-overflow: wrap
    footer: Introduction to SEM (2024)
    width: 1200
    height: 1000
editor:
  render-on-save: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(DiagrammeR)
library(ggplot2)
```

## Schedule & content {.smaller}

-   *Day 1*:
    -  General introduction to SEM to model ecological systems (lecture)
    -  Fitting SEMs to data (live demo)
    -  Model evaluation and reporting (exercises)
    -  Work on your research question /  SEM model (self-study)
-   *Day 2*: 
    -  Constructs: latent and composite variables (lecture)
    -  Fitting SEMs to data (live demo & exercises)
    -  Model visualization (lecture)
    -  Work on your research question /  SEM model (self-study)
-   *Day 3*:
    -  Work on your research question /  SEM model (self-study)
    -  Discussion with instructor and peers


## Motivation for SEM modelling

::: incremental
- System thinking to understand the direct and indirect cause-effect pathways in a system
- Consideration of relevant theory via the meta-model
- Theory is often more general than particular proxies used (e.g. climate change $\neq$ temperature change)
- How to connect particular proxies with general concepts?
:::


## Motivation for SEM modelling {.smaller}


```{dot}
//| fig-width: 9
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = LR, ranksep=1.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  
   subgraph cluster_bio {
        label="Biodiversity"
        fontname = Helvetica
		    color=lightgrey;
        evenness, richness;
        }


   subgraph cluster_env {
        label="Environment"
        fontname = Helvetica
     		color=lightgrey;
        nutrients;
        disturbance;
        }

    nutrients->richness;
    nutrients->evenness;
    nutrients->biomass;
    disturbance -> biomass;

   subgraph cluster_prod {
     label="Function";
     fontname = Helvetica;
     color=lightgrey;
     biomass; 
     }

    evenness->biomass ;
    richness->biomass ;
  
}
```

::: incremental
-   So far, only measured variables were considered (i.e. path modelling).
- However:
  -   Some components of a meta-model may be abstract and difficult to quantify and measure directly.
  -   Some components may be convenient to summarize as a single cause.
-   SEM provides constructs to address these situations: 
    1) latent variables 
    2) composite variables
:::

## Latent variables

```{dot}
//| fig-width: 6

digraph  {

  graph [rankdir = LR]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> Z;
    Z;
    
  {rank = max; Z}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```

::: incremental
-   Latent variables are unobserved.
-   In diagrams, latent variables are represented by oval node shape.\
-   Their influence is captured by one or more indicator variable(s).\
-   Direction of causality flows from latent variable to the observed variable(s) (reflective indicators).
:::

## Latent variables

```{dot}
//| fig-width: 6

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> Z;
    Z;
    
  {rank = max; Z}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```

::: incremental
-   Indicator variables "emerge" from the latent variable.\
-   Latent variables free of random or systematic measurement errors (unlike their indicators).
-   Identificability issues with two indicator variables require additional assumptions (e.g. similar effect of indicators on latent).
:::

## Latent variables - a word of caution
::: incremental
- The design of latent variables requires deep thinking and justification
- Just because you name a particular latent in a certain way, does not mean you capture the underling process (a.k.a. "naming fallacy")
- You have to justify the use of particular indicators to define a latent
:::

## Latent variables

Latents fitted in two steps:

. . .

1)  Confirmatory factor analysis
    -   Precursor to evaluation of any structural models in which the latent variables appear.
    -   All indicators need to be positively correlated to the latent variable (i.e., driver).
    -   For multi-indicator latent variables: check that the same process generated multiple indicators.

. . .

2)  Full model including latent variable
    -   To estimate effect of latent variable on other variables.

## Latent variables to model abstract concept

```{dot}
//| fig-width: 4

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="mass"];
   X2 [label="length"];
   X3 [label="width"];

   
   node [shape=oval]
    Y [label="body size"]
    
    Y -> X1;
    Y -> X2;
    Y -> X3;
    
  {rank = min; X1}
  {rank = same; X1;X2; X3}
}
```

Example body size

::: incremental
  - multiple indicators of body size: mass, length, width
  - check for positive correlation among indicators
  - With three indicator variable, coefficient of first indicator set to 1 by lavaan; use standardized effect for interpretation.
:::


## Latents to adjust for measurement error in predictors

```{dot}
//| fig-width: 6

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X [label="X with error"];
   Y [label="Y"];

    
    X -> Y;

    
}
```

::: incremental
-   Predictors are often measured with error.\
-   Regression assume that predictors are measured without error.\
-   What difference does it make?
:::

## Latents to adjust for measurement error in predictors

```{r}
#| echo: false
library(ggplot2); set.seed(1981981)
x <- rnorm(50); x2 <- x + rnorm(50); y <- 2*x; resid <- x-x2;
ggplot(data=NULL) + geom_point(aes(x=x2,y=y), colour="blue") +  
geom_segment(aes(x=x, xend = x2, y=y, yend = y), linetype="dashed", colour="red")+
stat_smooth( aes(x=x2,y=y), method="lm", se=F, colour="blue") + 
geom_abline(intercept=0,slope=2, colour="red")
```
Red = true slope, blue = OLS slope.

::: incremental
-   Measurement error biases the slope towards zero = regression dilution.
-   How to account for regression dilution?
:::

## Latents to adjust for measurement error in predictors

 

```{dot}
//| fig-width: 6

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="measurements 1"];
   X2 [label="measurement 2"];
    
   node [shape=oval]
   Z [label="latent"]
    
   Z -> X1;
   Z -> X2;
   Z -> Y;
       
  {rank = max; Y}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```


::: incremental
-   Possible to account for measurement error when multiple measurements were made.
-   Latent variable captures the underlying process giving rise to the indicator variables.
-   Estimating relationship between latent and response will lead to better slope estimate and higher R^2^.
:::

## Composite variables

::: incremental
- Composites are very flexible constructs.
- Allow to model interactions and nonlinear relationships.
- Allow to summarize multiple drivers into a single cause.
- Composites are shown as hexangular shapes in causal diagrams.
:::


## Composite variables

```{dot}
//| fig-width: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  X1 -> Y;
  X2 -> Y;
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Composites specify the collective influence of a set of variables.
- Here, land use is defined as the combination of disturbance and nutrients.
:::



## Composite variables

```{dot}
//| fig-width: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  node [shape=plain, fontname = Helvetica]
  E [label="0"]
  E -> Y [constraint=false];

  X1 -> Y;
  X2 -> Y;
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
-  Arrows point from the indicators to the composite (i.e., formative or causal indicators).
-  Values are determined by causes (i.e., indicators), thus error variance is set to 0.\
- Technically, composites are the predictions from linear combinations of variables.
- Estimation can cause convergence issues, lavaan can only handle a single composite per model.
- Solution: hand-code composites separately.
- Only standardised coefficients of composites are interpretable.
:::


## Using composites to model nonlinear effects{.smaller}


```{dot}
//| fig-width: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="nutrients"];
   X2 [label=<nutrients<SUP>2</SUP>>];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="nutrient effect"]

  X1 -> Y;
  X2 -> Y;
  Y -> Z;

  X1:w -> X2:w [dir="both"]

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Nutrient effects on biomass may be nonlinear (e.g. hump-shaped)
- Can be modelled by adding a linear and a squared term of nutrients
- Composite to capture this joint effect of nutrient addition
- Check validity of nonlinear effect by only constructing composite first (linear model)
- Use standardized coefficients for interpretation
:::

## Distinguishing latent and composite variables 

Pay attention to the flow of causation!

.  .  .

1) Latents
  - From construct to indicators, i.e., indicators are driven by an underlying, unmeasured process.
  - Indicators need to be correlated to each other, since a latent represents an underlying, data-generating process.\

. . .  

2) Composites 
  - flow of causation is reversed and the indicators are independent entities.\
  - There is no assumption about the relation between the indicators (can be positive/negative).


## Distinguishing latent and composite variables 

Rules of thumb:

::: incremental
1) Latent variable -> indicators are redundant (e.g. multiple measurements)
2) Composite variable -> the meaning of the construct changes after dropping one of the indicators
:::


## Exercise:

-   Revisit the meta-model you have drawn yesterday:
    -   Which contructs could be modelled with a latent or composite variable?
    -   Adapt your meta-model accordingly.
- Revisit the meta-model of your dataset:
  - Are there concepts that could be modelled with a latent or composite variable?


## Interactions {.smaller}

```{dot}
//| fig-width: 6
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X1 [label="nutrients"]; 
  X2 [label="disturbance"]; 
  Y [label="biomass"];
  middleman

  # several 'edge' statements
    X1->middleman[arrowhead=none];
    middleman->Y ;
    middleman[shape=none width=0 height=0 label="" style=invis]
    X2->middleman [constraint=false, arrowhead=diamond];

}
```

::: incremental
-   In nature, things often are contingent on each other.
-   For instance, the effect of nutrients on plant growth, may depend on how disturbed the environment is.\
-   Moderation: effects are different when combined.\
-   In regression models, the coefficient for the interaction estimates the effect of the product of the two predictors.\
-   The coefficient of the interaction describes the deviation from additivity (can be positive or negative).
:::

## Interactions 

Interactions can be modelled in different ways in lavaan:

1)  Composites
2)  Multiple groups

## Interactions (composite)
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        richness;
        nutrients;
        disturbance;
        middleman

  middleman[shape=none width=0 height=0 label="" style=invis]
disturbance->middleman [constraint=false, arrowhead=diamond];

    middleman->richness;

    richness->biomass;

    nutrients -> middleman[arrowhead=none];
     
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

- Interactions can also be modelled with composites.
- Preferable if only one or few interactions are considered.


## Interactions (multiple groups)

::: {.column width="40%"}
Group = Disturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness [style="dashed"];
    nutrients -> evenness;

    evenness-> biomass;
    richness->biomass [style="dashed"];

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```
:::

::: {.column width="40%"}
Group = Undisturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness [style="dashed"];
    nutrients -> biomass [color="red"];

    evenness-> biomass [style="dashed"];
    richness->biomass;

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

:::


::: incremental
-   Multigroup fitting allows all coefficients to vary among groups.
-   Groups have to be categorical (e.g., treatment, age class).
-   Test interaction of group factor with many or all paths in the causal diagram.
- Statistical comparison of two causal diagrams $\neq$ fitting two separate SEMs and compare coefficients.
:::

## Interactions (multiple groups)

Lavaan allows to introduce equality constraints on various aspects via the `group.equal` argument:

``` r
sem(model, group = "age_class", group.equal = c("regressions"), data = dd)
```

. . .

Adding constraints via `group.equal`:

``` r
group.equal=c("intercepts", "means", "regressions", "residuals")
```

## Interactions (multiple groups)

Even more control by having the same name for different parameters:

``` r
model <- ' 
y ~ c("b1", "b1") * x1 + c("b2", "b2") * x3
x2 ~ c("b3", "b4") * x1
x3 ~ c("b5", "b5") * x2'
```

Same coefficients for all but the effect of x1 on x2.




<!-- # Complex sampling structure

## Complex sampling structure

::: incremental
-   Needed when data is nested, e.g.,
    -   within sites.
    -   within groups such as families, nests.
-   Nesting violates the principle of independent and identically distributed observations.\
-   Necessary to account for data structure in the model.
:::

## Complex sampling structure

``` r
library("lavaan.survey")
```

-   The add-on package lavaan.survey allows the analysis of stratified, clustered or weighted data.

. . .

-   Lavaan objects are processed with a specific data structure:
    1)  Initialize the design\
    2)  Post-process the lavaan object and compute the adjusted results.
    3)  Corrected lavaan object as result.

## Complex sampling structure

``` r
design <- svydesign(ids = ~ plot, strata = ~ field, 
nest = TRUE, data = dat)
summary(design)

fit.nested <- lavaan.survey(lavaan.fit = model, 
survey.design = design)
```

-   Needed to specify the study design
-   Here we have plots (`ids`) nested in fields (`strata`)
-   Next, we can refit the `simple` model from before with `lavaan.survey` using the specified study design as an argument.
 -->

 
<!-- # Spatial autocorrelation -->

<!-- ## Spatial autocorrelation -->

<!-- - Why is autocorrelation problematic? -->

<!-- - Autocorrelation is a sign of non-independence, violating assumptions of linear models that data points are independent. -->

<!-- - Does not influence estimates, but their standard errors and p values. -->

<!-- - You think you have more information than you actually have. -->

# Questions?

# Live coding session

# Your turn: working with the Seabloom dataset

## Exercise 1

Start with the following model:

``` r
library("lavaan")

simple <-
"mass.above ~ nadd + disk + rich + even
rich ~ nadd 
even ~ nadd

rich ~~ even"

fit.simple <- sem(simple, data = seabloom, estimator = "MLM")
summary(fit.simple)
```

## Exercise 2

```{dot}
//| fig-width: 4

digraph  {

  graph [rankdir = TB, ranksep=.5]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   X3 [label="ens.pie"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> X3;
    
  {rank = min; X1;X2; X3}
}
```


-   Construct latent variable diversity based on richness, evenness and ens.pie:
    -   Run confirmatory factor analysis
    -   What do you conclude?

## Exercise 3

```{dot}
//| fig-width: 6
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

    node [shape=oval, fontname = Helvetica]
    diversity

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;

    disturbance->diversity;
    nutrients->diversity;
    disturbance -> biomass;
    diversity->biomass ;

    diversity -> evenness;
    diversity -> richness;

 { rank="max"; nutrients; disturbance}

 { rank="min"; evenness; richness}
}
```

-   Construct latent variable diversity only based on richness and evenness:
    -   Run confirmatory factor analysis
    -   Include latent variable into full model
    -   What do you conclude?

## Exercise 4

```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

    node [shape=hexagon, fontname = Helvetica]
    landuse

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;

    landuse->richness;
    landuse -> evenness;
    landuse -> biomass;

    evenness-> biomass;
    richness->biomass;

    nutrients -> landuse;
    disturbance -> landuse;
  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```


-   Construct composite variable landuse based on nutrients and disturbance (disk):
    -   What do you conclude?
    -   Build composite manually.

## Exercise 5

```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        middleman

  middleman[shape=none width=0 height=0 label="" style=invis]
disturbance->middleman [constraint=false, arrowhead=diamond];

    middleman->richness;
    middleman -> evenness;

    evenness-> biomass;
    richness->biomass;

    nutrients -> middleman[arrowhead=none];
     
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```



-   Investigate possible interaction between disturbance and nutrient addition on AGB
    -   Construct composite to model an interaction
    -   Incorporate composite into full model
- What do you conclude?


## Exercise 6



::: {.column width="40%"}
Group = Disturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness;
    nutrients -> biomass;

    evenness-> biomass;
    richness->biomass;

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```
:::

::: {.column width="40%"}
Group = Undisturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness;
    nutrients -> biomass;

    evenness-> biomass;
    richness->biomass;

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

:::


-   Investigate possible interaction between disturbance and nutrient addition on AGB:
    -   Use multi-group fitting to explore interaction.
-   What do you conclude?

<!-- ## Exercise 7

-   Account for nested experimental design with the `lavaan.survey` package:
    -   Add individual plots nested in fields.
 -->