---
author: "Frank Pennekamp"
title: "Introduction to structural equation modelling"
subtitle: "Day 2: Advanced modelling"
institute: 
- Department of Evolutionary Biology and Environmental Sciences, University of Zurich
date: "9.11.2022"
format:
  revealjs: 
    smaller: false
    scrollable: false
    theme: simple
    slide-number: true
    show-slide-number: all
    code-overflow: wrap
    footer: Introduction to SEM (2022)
    width: 1200
    height: 800
editor:
  render-on-save: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(DiagrammeR)
library(ggplot2)
```

## Schedule & content {.smaller}

-   *Day 1*:
    -   General introduction to SEM to model ecological systems
    -   Fitting SEMs to data (live demo)
    -   Model evaluation and reporting (exercises)
    -   Discussion with James Grace
-   *Day 2*: 
    -   Latent and composite variables, interactions, complex sampling designs
    -   Fitting SEMs to data (live demo & exercises)
    -   Model visualization
    -   Discussion with James Grace
-   *Day 3*:
    -   Self-study with possibility to meet with instructor(s)

## Revisiting the meta-model {.smaller}

```{dot}
//| fig-width: 6
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, splines = curved, rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  
   subgraph cluster_bio {
        label="Biodiversity"
        fontname = Helvetica
		    color=lightgrey;
        evenness, richness;
        }


   subgraph cluster_env {
        label="Environment"
        fontname = Helvetica
     		color=lightgrey;

        nutrients;
        disturbance;
        }

    nutrients->richness;
    nutrients->evenness;
    nutrients->biomass;
    disturbance -> biomass;

   subgraph cluster_prod {
     label="Productivity";
     fontname = Helvetica;
     color=lightgrey;
     biomass; 
     }

    evenness->biomass ;
    richness->biomass ;
  
}
```

::: incremental
-   Meta-model = conceptual model to link data with theory.
-   So far, only manifest (measured) variables were considered.\
-   However, some aspects of a meta-model may be difficult to quantify and measure directly.\
-   This is because they represent an abstract, multifaceted concept (e.g., "biodiversity").
-   Latent and composite variables can help with these situations.\
:::

# Latent variables

## Latent variables

```{dot}
//| fig-width: 6

digraph  {

  graph [rankdir = LR]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> Z;
    Z;
    
  {rank = max; Z}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```

::: incremental
-   Latent variables are unobserved.
-   Bridge between observed data and theory.
-   Their influence is captured by one or more indicator variable(s).\
-   Direction of causality flows from latent variables to the observed variables (reflective indicators).
-   Graphically, latent variables are represented by oval node shape.\
:::

## Latent variables

```{dot}
//| fig-width: 6

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> Z;
    Z;
    
  {rank = max; Z}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```

::: incremental
-   Indicator variables "emerge" from the latent variable.\
-   Latent variables free of random or systematic measurement errors (unlike their indicators).
-   All indicators should be positively correlated to the latent variable (i.e., driver).
:::

## Latent variables

::: incremental
-   Latents can be both endogenous and exogenous.
-   *Measurement model* focuses solely on relating indicators to latent variables.
-   *Structural model* is one with directed paths between latent variables.
:::

## Latent variables

Latents often fitted in two steps:

. . .

1)  Confirmatory factor analysis

-   Multi-indicator latent variables: the same process generated multiple indicator variables.\
-   Precursor to evaluation of any structural models in which the latent variables appear.

. . .

2)  Full model including latent variable

-   To estimate effect of latent variable on other variables.

## Using latents to adjust for measurement error in the explanatory variables

::: incremental
-   Variables are often measured with error.\
-   Regression approaches assume that predictor variables are measured without error.\
-   What difference does it make?
:::

## Regression dilution {.smaller}

```{r}
#| echo: false
library(ggplot2); set.seed(1981981)
x <- rnorm(50); x2 <- x + rnorm(50); y <- 2*x; 
ggplot(data=NULL) + geom_point(aes(x=x2,y=y), colour="blue") +   
stat_smooth( aes(x=x2,y=y), method="lm", se=F, colour="blue") + 
geom_abline(intercept=0,slope=2, colour="red")
```

::: incremental
-   Red = true slope, blue = OLS slope.
-   Measurement error biases the slope towards zero.
-   How to account for regression dilution?
:::

## Latents to adjust for noise in the explanatory variables

::: incremental
-   Possible to account for measurement error when multiple measurements were made.
-   Latent variable captures the underlying process giving rise to the indicator variables.
-   Estimating relationship between latent and response will lead to better slope estimate and higher R^2^.
:::

# Composite variables

## Composite variables {.smaller}

```{dot}
//| fig-width: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  node [shape=plain, fontname = Helvetica]
  E [label="0"]
  E -> Y [constraint=false];

  X1 -> Y;
  X2 -> Y;
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
-   Composites to specify the collective influence of a set of variables (e.g., land use).
-   Composites arise from the indicators (in contrast to latents).\
-   Values are determined by causes (i.e., indicators), thus error variance is set to 0.\
-   Arrows point from the indicators to the composite (i.e., formative or causal indicators).
-   Composites are shown as hexangular shapes.
:::

## Composite variables

```{dot}
//| fig-width: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  node [shape=plain, fontname = Helvetica]
  E [label="0"]
  E -> Y [constraint=false];

  X1 -> Y;
  X2 -> Y;
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Composites are very flexible constructs
- Allow to model interactions and nonlinear relationships
- Technically they are predictions
:::

## Key differences between latent and composite variables {.smaller}

::: columns
::: {.column width="60%"}
::: incremental
-   Flow of causation:
    -   *Latents*: from construct to indicators, i.e., indicators are driven by an underlying, unmeasured process.
    -   Indicators need to be correlated to each other, since a latent represents an underlying, data-generating process.\
    -   *Composites*: flow of causation is reversed and the indicators are independent entities.\
    -   There is no assumption about the relation between the indicators (can be positive/negative).
:::
:::

::: {.column width="40%"}
```{dot}
//| fig-width: 4

digraph boxes_and_circles {

  graph [splines = curved, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];
   X3 [label="disturbance"];
   X4 [label="nutrients"];

   
   node [shape=oval, fontname = Helvetica]
   Y1 [label="diversity"]
    
   Y1 -> X1 [dir=back];
    Y1 -> X2 [dir=back];
    Y1 -> Z;
     
  node [shape=hexagon, fontname = Helvetica]
  Y2 [label="land use"]

  X3 -> Y2;
  X4 -> Y2;
  Y2 -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1; X2; X3; X4}
  {rank = "same"; Y1; Y2;}
}
```
:::
:::

## Key differences between latent and composite variables

How distinguish latent and composites:

::: incremental
-   Rule of thumb 1: if the indicators are redundant, they likely belong to a latent variable.\
-   Rule of thumb 2: If the meaning of the construct changes after dropping one of the indicators, the construct likely is a composite.\
:::

# Exercise

## Exercise:

-   Revisit the meta-model you have drawn yesterday:
    -   Which contructs could be modelled with a latent or composite variable?
    -   Adapt your meta-model accordingly.

# Interactions

## Interactions {.smaller}

```{dot}
//| fig-width: 6
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X1 [label="nutrients"]; 
  X2 [label="disturbance"]; 
  Y [label="biomass"];
  middleman

  # several 'edge' statements
    X1->middleman[arrowhead=none];
    middleman->Y ;
    middleman[shape=none width=0 height=0 label="" style=invis]
    X2->middleman [constraint=false, arrowhead=diamond];

}
```

::: incremental
-   In nature, things often are contingent on each other.
-   For instance, the effect of nutrients on plant growth, may depend on how disturbed the environment is.\
-   Moderation: effects are different when combined.\
-   In regression, the interaction coefficient estimates the effect of the product of the two predictors.\
-   Interaction coefficient estimates the deviation from additivity.
:::

## Interactions

Interactions can be modelled in different ways in lavaan:

1)  Multiple groups
2)  Composites

## Interactions (multi group)

``` r
sem(model, group = "age_class", data = dd)
```

::: incremental
-   Multigroup fitting allows coefficients to vary among groups.\
-   Lavaan offers the "group" argument to specify for which groups coefficients should be estimated.\
-   Importantly, groups have to be categorical (e.g., sex, age class).
:::

## Interactions (multi group)

Lavaan allows to introduce equality constraints on various aspects via the `group.equal` argument:

``` r
sem(model, group = "age_class", group.equal = c("regressions"), data = dd)
```

. . .

Adding constraints via `group.equal`:

``` r
group.equal=c("intercepts", "means", "regressions", "residuals")
```

## Interactions (multi group)

Even more control by having the same name for different parameters:

``` r
model <- ' 
y ~ c("b1", "b1") * x1 + c("b2", "b2") * x3
x2 ~ c("b3", "b4") * x1
x3 ~ c("b5", "b5") * x2'
```

Same coefficients for all but the effect of x1 on x2.

# Complex sampling structure

## Complex sampling structure

::: incremental
-   Needed when data is nested, e.g.,
    -   within sites.
    -   within groups such as families, nests.
-   Nesting violates the principle of independent and identically distributed observations.\
-   Necessary to account for data structure in the model.
:::

## Complex sampling structure

``` r
library("lavaan.survey")
```

-   The add-on package lavaan.survey allows the analysis of stratified, clustered or weighted data.

. . .

-   Lavaan objects are processed with a specific data structure:
    1)  Initialize the design\
    2)  Post-process the lavaan object and compute the adjusted results.
    3)  Corrected lavaan object as result.

## Complex sampling structure

``` r
design <- svydesign(ids = ~ plot, strata = ~ field, 
nest = TRUE, data = dat)
summary(design)

fit.nested <- lavaan.survey(lavaan.fit = model, 
survey.design = design)
```

-   Needed to specify the study design
-   Here we have plots (`ids`) nested in fields (`strata`)
-   Next, we can refit the `simple` model from before with `lavaan.survey` using the specified study design as an argument.

<!-- # Spatial autocorrelation -->

<!-- ## Spatial autocorrelation -->

<!-- - Why is autocorrelation problematic? -->

<!-- - Autocorrelation is a sign of non-independence, violating assumptions of linear models that data points are independent. -->

<!-- - Does not influence estimates, but their standard errors and p values. -->

<!-- - You think you have more information than you actually have. -->

# Questions?

# Live coding session

# Your turn: working with the Seabloom dataset

## Exercise 1

Start with the following model:

``` r
library("lavaan")

simple <-
"mass.above ~ nadd + disk + rich + even + precip.mm
rich ~ nadd + precip.mm
even ~ nadd + precip.mm

rich ~~ even"

fit.simple <- sem(simple, data = seabloom, estimator = "MLM")
summary(fit.simple)
```

## Exercise 2

```{dot}
//| fig-width: 4

digraph  {

  graph [rankdir = TB, ranksep=.5]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   X3 [label="ens.pie"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1;
    Y -> X2;
    Y -> X3;
    
  {rank = min; X1;X2; X3}
}
```


-   Construct latent variable diversity based on richness, evenness and ens.pie:
    -   Run confirmatory factor analysis
    -   What do you conclude?

## Exercise 3

```{dot}
//| fig-width: 6
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5, splines=curved]

    node [shape=oval, fontname = Helvetica]
    diversity

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        precipitation;

    disturbance->diversity;
    nutrients->diversity;
    disturbance -> biomass;
    precipitation -> biomass;
    diversity->biomass ;

    diversity -> evenness;
    diversity -> richness;
  
 { rank="same"; nutrients; disturbance, precipitation}
}
```

-   Construct latent variable diversity only based on richness and evenness:
    -   Run confirmatory factor analysis
    -   Include latent variable into full model
    -   What do you conclude?

## Exercise 4

```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5, splines=curved]

    node [shape=oval, fontname = Helvetica]
    diversity

    node [shape=hexagon, fontname = Helvetica]
    landuse

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        precipitation;

    landuse->diversity;
    landuse -> biomass;
    precipitation -> biomass;
    diversity->biomass ;

    nutrients -> landuse;
    disturbance -> landuse;

    diversity -> evenness;
    diversity -> richness;
  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```


-   Construct composite variable landuse based on nutrients and disturbance (disk):
    -   What do you conclude?
    -   Build composite manually.

## Exercise 5

-   Investigate possible interaction between disturbance and nutrient addition on AGB
    -   Construct composite to model an interaction
    -   Incorporate composite into full model
- What do you conclude?


## Exercise 6

-   Investigate possible interaction between disturbance and nutrient addition on AGB:
    -   Use multi-group fitting to explore interaction.
-   What do you conclude?

## Exercise 7

-   Account for nested experimental design with the `lavaan.survey` package:
    -   Add individual plots nested in fields.
