---
author: "Frank Pennekamp"
title: "Introduction to path and structural equation modelling"
subtitle: "Day 2: Structural equation modelling"
institute: 
- Department of Evolutionary Biology and Environmental Sciences, University of Zurich
date: "11/6/2025"
date-format: long
format:
  revealjs: 
    smaller: false
    scrollable: false
    theme: simple
    slide-number: true
    show-slide-number: all
    code-overflow: wrap
    footer: Introduction to SEM (2025)
    width: 1200
    height: 1000
editor:
  render-on-save: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(DiagrammeR)
library(ggplot2)
```

## Schedule & content {.smaller}

-   *Day 1*:
    -  General introduction to path modelling for ecological systems (lecture)
    -  Fitting path models to data (live demo)
    -  Case study: fitting path models to data (exercises)
    -  Work on your research question /  meta-model (self-study)
-   *Day 2*: 
    -  Constructs: latent and composite variables (lecture)
    -  Fitting SE models to data (live demo & case study exercises)
    -  Work on your research question /  SE model (self-study)
    -  Progress update
-   *Day 3*:
    -  Model visualization (lecture)
    -  Work on your research question /  SE model (self-study)
    -  Presentation of your SE model and discussion with instructor and peers

## SE modelling for causal understanding of systems

::: incremental
- Meta-model aims to connect general theory with study specific model and data
- Theory is often more general than particular proxies used (e.g. climate change includes temperature mean, extremes, precipitation changes etc.)
- Constructs can bridge between general theory and the complexity of case studies:
  - Capture underlying processes and multiple causes
  - Interactions between variables
  - Nonlinear relationships
:::

## Motivation for SE modelling {.smaller}


```{dot}
//| fig-width: 9
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = LR, ranksep=1.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  
   subgraph cluster_bio {
        label="Biodiversity"
        fontname = Helvetica
		    color=lightgrey;
        evenness, richness;
        }


   subgraph cluster_env {
        label="Human impact"
        fontname = Helvetica
     		color=lightgrey;
        nutrients;
        disturbance;
        }

    nutrients->richness;
    nutrients->evenness;
    nutrients->biomass;
    disturbance -> biomass;

   subgraph cluster_prod {
     label="Function";
     fontname = Helvetica;
     color=lightgrey;
     biomass; 
     }

    evenness->biomass ;
    richness->biomass ;
  
}
```

::: incremental
- So far, only observed variables were considered (i.e. path model).
- However, many ecological concepts are:
  -  abstract and difficult to quantify and measure directly
  -  multifaceted and multidimensional
-  Latent and composite variables help to capture such concepts
:::

# Latent variables

## Latent variables in the social sciences and psychology

> Instrument development: The process of creating a new instrument [a set of specific measurements] for measuring an unobserved or latent construct, such as depression, sociability, or fourth-grade mathematics ability.

::: incremental
- In the social sciences, latent variables are widely used to capture abstract concepts.
- Human attitudes and motivations are assumed to be “deeply latent” and only discernible indirectly.
- Moreover, the properties often of primary interest itself and used as instruments.
- Powerful but currently underused tool in ecology.
:::


## Latent variables {.smaller}

```{dot}
//| fig-width: 6
//| fig-height=: 1
digraph  {

  graph [rankdir = LR]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   Z [label="biomass"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1 [style="dashed"];
    Y -> X2 [style="dashed"];
    Y -> Z;
    Z;

  
  {rank = max; Z}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```

::: incremental
-   Latent variables represented by oval node shape.\
-   The latent variable itself is unobserved, its influence is captured by indicator variable(s).\
-   Latents capture the common variance or overlapping information for a set of measures. 
-   Direction of causality flows from latent variable to the (reflective) indicator(s).
-   Latent variables are free of random or systematic measurement errors (unlike their indicators).
:::

## Fitting latent variables

::: incremental
- Latent variable can have 1 to many indicators:
  - 1 indicator: 
    - assumes measured variable is perfect representation of the latent. 
    - makes a conceptual distinction between the concept of interest and the observed measure.
  - 2 indicators:
    - requires assumptions to identify the model. 
    - possible to remove measurement error.
  - 3 or more indicators: 
    - allows to test whether or not the same process generated the indicators.
:::

## Fitting latent variables

Latents fitted in two steps:

. . .

1)  Measurement model (Confirmatory Factor Analysis)
    -   Precursor to any SEM in which the latent variables appear.
    -   For multi-indicator latent variables: check that the same process generated multiple indicators.
    -   All indicators need to be correlated with similar loadings to justify an underlying latent variable (i.e., the cause).

. . .

2)  Structural model
    -   Estimates relationships betweenmultiple latent variables and between latent and other variables.

## Latent example: measurement error in predictor

```{dot}
//| fig-width: 8
//| fig-height: 5


digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X [label="X with error"];
   Y [label="Y"];

    
    X -> Y;

    
}
```

::: incremental
-   Predictors are often measured with error.\
-   Regression assumes no error in X.\
-   What if this assumption is violated?
:::

## Latent example: measurement error in predictor

```{r}
#| echo: false
library(ggplot2); set.seed(1981981)
x <- rnorm(50); x2 <- x + rnorm(50); y <- 2*x; resid <- x-x2;
ggplot(data=NULL) + geom_point(aes(x=x2,y=y), colour="blue") +  
geom_segment(aes(x=x, xend = x2, y=y, yend = y), linetype="dashed", colour="red")+
stat_smooth( aes(x=x2,y=y), method="lm", se=F, colour="blue") + 
geom_abline(intercept=0,slope=2, colour="red")
```
Red = true slope, blue = OLS slope.

::: incremental
-   Measurement error biases the slope towards zero = regression dilution.
-   How to account for regression dilution?
:::

## Latent example 1: measurement error in predictor

```{dot}
//| fig-width: 8
//| fig-height: 5

digraph  {

    graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="1st noisy measurement"];
   X2 [label="2nd noisy measurement"];
    
   node [shape=oval]
   Z [label="underlying process"]
    
   Z -> X1 [style="dashed"];
   Z -> X2 [style="dashed"];
   Z -> Y;
       
  {rank = max; Y}
  {rank = min; X1}
  {rank = same; X1;X2}
}
```


::: incremental
-   Latent captures the overlapping information in the repeated measures.
-   Better slope estimate and higher R^2^ when regressing Y on latent.
:::

## Latent example 2: body size {.smaller}

```{dot}
//| fig-width: 4
//| fig-height=: 2
digraph  {

  graph [rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="mass"];
   X2 [label="length"];
   X3 [label="width"];
   X4 [label="height"];

   
   node [shape=oval]
  Y [label="body size"]
  
  Y -> X1 [style="dashed"];
  Y -> X2[style="dashed"];
  Y -> X3[style="dashed"];
  Y -> X4[style="dashed"];
   
  {rank = max; Y}
  {rank = min; X1}
  {rank = same; X1;X2; X3;X4}
}
```

::: incremental
- Body size is multidimensional.
  - Multiple indicators of body size: mass, length, width
  - Identifiability: 
    - Loading of first indicator set to 1 by lavaan (sets scale of latent to 1st indicator).
    - Because of this, use standardized effect for interpretation.
    - Latents with two indicators require additional assumptions (e.g. similar loadings of latent on indicators).
:::

## Latent variables - a word of caution
::: incremental
- The design of latent variables requires deep thinking and justification
- The choice of particular indicators to define a latent variable needs justification.
- The naming fallacy: naming a latent variable in a certain way does not guarantee you capture the underling process.
:::

# Composite variables

## Composite variables

::: incremental
- Composites can be used flexibly and require less assumptions than latents:
  - Summarize multiple causes under a single composite (dimension reduction).
  - Allow to model interactions and nonlinear relationships.
- Composites are shown as hexangular shapes in causal diagrams.
:::


## Example composite: joint cause landuse

```{dot}
//| fig-width: 8
//| fig-height: 5


digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  node [shape=plain, fontname = Helvetica]
  E [label="0"]
  E -> Y [constraint=false];

  X1 -> Y [style="dashed"];
  X2 -> Y [style="dashed"];
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Composites specify the collective influence of a set of variables.
- Arrows point from the indicators to the composite (i.e., formative or causal indicators).
- The values of the composite determined by causes (i.e., indicators), thus error variance is set to 0.\
  - Technically, predictions from linear combinations of variables.
:::

## Example composite: joint cause landuse

```{dot}
//| fig-width: 8
//| fig-height: 5

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="disturbance"];
   X2 [label="nutrients"];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="land use"]

  node [shape=plain, fontname = Helvetica]
  E [label="0"]
  E -> Y [constraint=false];

  X1 -> Y [style="dashed"];
  X2 -> Y [style="dashed"];
  Y -> Z;

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Estimation can cause convergence issues, lavaan can only handle a single composite per model.
  - Solution: hand-code composites.
- Only standardised coefficients interpretable.
:::

## Example composite: model nonlinear effects


```{dot}
//| fig-width: 8
//| fig-height: 2

digraph boxes_and_circles {
   
   graph [overlap = false, fontsize = 10, rankdir = LR, fontname = Helvetica]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="nutrients"];
   X2 [label=<nutrients<SUP>2</SUP>>];
   Z [label="biomass"];

   
   node [shape=hexagon, fontname = Helvetica]
  Y [label="nutrient effect"]

  X1 -> Y [style="dashed"];
  X2 -> Y [style="dashed"];
  Y -> Z;

  X1:w -> X2:w [dir="both"]

  {rank = "max"; Z}
  {rank = "same"; X1;X2;}
}
```

::: incremental
- Nutrient effects on biomass may be nonlinear (e.g. hump-shaped)
- Can be modelled by adding a linear and a squared term of nutrients
- Check validity of nonlinear effect by only constructing composite first (linear model)
:::

<!--- ## Distinguishing latent and composite variables 

Pay attention to the flow of causation!

.  .  .

1) Latents
  - From construct to indicators, i.e., indicators are driven by an underlying, unmeasured process.
  - Indicators need to be positively correlated, since a latent represents an underlying, data-generating process.\

. . .  

2) Composites 
  - flow of causation is reversed and the indicators are independent entities.\
  - There is no assumption about the relation between the indicators (can be positive/negative).

-->


## Distinguishing latent and composite variables 

Rules of thumb:

::: incremental
- Latent: 
  - indicators are redundant, e.g. 
    - multiple measurements of the same process
    - multiple proxies of the same process
- Composite:
  - the meaning of the construct changes after dropping one of the indicators
:::


## Exercise:

- Revisit your meta-model:
  - Are there concepts that could be modelled with a latent or composite variable?
  - Adapt your meta-model accordingly.

# Interactions

## Interactions

```{dot}
//| fig-width: 6
//| fig-height: 3
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X1 [label="nutrients"]; 
  X2 [label="disturbance"]; 
  Y [label="biomass"];
  middleman

  # several 'edge' statements
    X1->middleman[arrowhead=none];
    middleman->Y ;
    middleman[shape=none width=0 height=0 label="" style=invis]
    X2->middleman [constraint=false, arrowhead=diamond];

}
```

::: incremental
-   Effect of one variable may depend on another variable.
-   E.g., nutrient effect on plant growth, may depend on whether the environment was disturbed in the past.\
-   The interaction coefficient captures possible positive or negative deviation from additivity (e.g., the sum of the two effects in isolation).
-   In a regression model, interaction coefficient estimates the effect of the product of the two predictors.\
:::



## Interactions 

Two ways to model interactions in lavaan:

1)  Interactions between **continous** variables: composite variable
2)  Interactions with **categorical** variables: multi-group SEM

## Interactions between continuous variables (composite)
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        richness;
        nutrients;
        disturbance;
        middleman

  middleman[shape=none width=0 height=0 label="" style=invis]
disturbance->middleman [constraint=false, arrowhead=diamond];

    middleman->richness;

    richness->biomass;

    nutrients -> middleman[arrowhead=none];
     
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

::: incremental
- Include interaction term in composite.
- Preferable when only a few interactions are considered.
:::

## Interactions with categorical variable (multi-group)

::: {.column width="40%"}
Group = Disturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness;

    evenness-> biomass [penwidth=3, color="red"];
    richness->biomass;

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```
:::

::: {.column width="40%"}
Group = Undisturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness [penwidth=3, color="red"];
   # nutrients -> biomass [color="red"];

    evenness-> biomass
    richness->biomass;

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

:::


::: incremental
- Multi-group SEM allows all coefficients to vary among groups.
- Groups have to be categorical (e.g., treatment, age class).
- Allows formal tests of equality among causal diagrams
:::

## Interactions with categorical variable (multi-group)

::: incremental
-   Test interaction of a categorical factor with many or all paths in the causal diagram.
- Lavaan allows to test various equality constraints:
   - Equal path coefficients, intercepts, residuals among groups
- Even more control by naming and fixing parameters
  - Coefficients with same name are estimated to be the same
  - Fixing coefficients to 0 allows graph to vary among groups.
::: 



<!-- ``` r
sem(model, group = "age_class", group.equal = c("regressions"), data = dd)
```

. . .

Adding constraints via `group.equal`:

``` r
group.equal=c("intercepts", "means", "regressions", "residuals")
```

## Interactions (multiple groups)

Even more control by having the same name for different parameters:

``` r
model <- ' 
y ~ c("b1", "b1") * x1 + c("b2", "b2") * x3
x2 ~ c("b3", "b4") * x1
x3 ~ c("b5", "b5") * x2'
```

Same coefficients for all but the effect of x1 on x2.




<!-- # Complex sampling structure

## Complex sampling structure

::: incremental
-   Needed when data is nested, e.g.,
    -   within sites.
    -   within groups such as families, nests.
-   Nesting violates the principle of independent and identically distributed observations.\
-   Necessary to account for data structure in the model.
:::

## Complex sampling structure

``` r
library("lavaan.survey")
```

-   The add-on package lavaan.survey allows the analysis of stratified, clustered or weighted data.

. . .

-   Lavaan objects are processed with a specific data structure:
    1)  Initialize the design\
    2)  Post-process the lavaan object and compute the adjusted results.
    3)  Corrected lavaan object as result.

## Complex sampling structure

``` r
design <- svydesign(ids = ~ plot, strata = ~ field, 
nest = TRUE, data = dat)
summary(design)

fit.nested <- lavaan.survey(lavaan.fit = model, 
survey.design = design)
```

-   Needed to specify the study design
-   Here we have plots (`ids`) nested in fields (`strata`)
-   Next, we can refit the `simple` model from before with `lavaan.survey` using the specified study design as an argument.
 -->

 
<!-- # Spatial autocorrelation -->

<!-- ## Spatial autocorrelation -->

<!-- - Why is autocorrelation problematic? -->

<!-- - Autocorrelation is a sign of non-independence, violating assumptions of linear models that data points are independent. -->

<!-- - Does not influence estimates, but their standard errors and p values. -->

<!-- - You think you have more information than you actually have. -->

# Questions?

# Live coding session

# Your turn: working with the Seabloom dataset

## Set up

Start with the following model:

``` r
library("lavaan")

simple <-
"mass.above ~ nadd + disk + rich + even
rich ~ nadd 
even ~ nadd

rich ~~ even"

fit.simple <- sem(simple, data = seabloom, estimator = "MLM")
summary(fit.simple)
```

## Exercise 1

```{dot}
//| fig-width: 4

digraph  {

  graph [rankdir = TB, ranksep=.5]
   
   node [shape=box, fontname = Helvetica]
   X1 [label="evenness"];
   X2 [label="richness"];
   X3 [label="ens.pie"];

   
   node [shape=oval]
    Y [label="diversity"]
    
    Y -> X1 [style="dashed"];
    Y -> X2 [style="dashed"];
    Y -> X3 [style="dashed"];
    
  {rank = min; X1;X2; X3}
}
```


-   Construct a latent variable called diversity using richness and evenness as indicators and ens.pie:
    - Use confirmatory factor analysis
    - Troubleshoot potential issues

## Exercise 2

```{dot}
//| fig-width: 6
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=1]

    node [shape=oval, fontname = Helvetica]
    diversity

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        biomass;

    disturbance->diversity;
    nutrients->diversity;
    disturbance -> biomass;
    nutrients->biomass;
    diversity->biomass ;

    diversity -> evenness [style="dashed"];
    diversity -> richness [style="dashed"];

 { rank="min"; evenness; richness}

 { rank="same"; nutrients; diversity; disturbance}

 { rank="max"; biomass}
 
 # Add invisible edges to force left-to-right ordering
 nutrients -> diversity [style=invis];
 diversity -> disturbance [style=invis];

}
```

-   Construct a latent variable called diversity using richness and evenness as indicators:
    -   Use confirmatory factor analysis to confirm validity of latent
    -   Include latent variable into structural equation model
-   What do you conclude?

## Exercise 3

```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

    node [shape=hexagon, fontname = Helvetica]
    landuse

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        biomass;

    nutrients->richness;
    nutrients -> evenness;

    landuse -> biomass;

    evenness-> biomass;
    richness->biomass;

    nutrients -> landuse [style="dashed"];
    disturbance -> landuse [style="dashed"];

    evenness-> richness [dir=both, minlen=3] ;

  
 { rank="same"; evenness, richness}
}
```


-   Construct composite variable landuse based on nutrients and disturbance in lavaan.
-   Construct composite manually.
- Compare original model with this one. Which one is better supported?

## Exercise 4

```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;
        disturbance;
        middleman

  middleman[shape=none width=0 height=0 label="" style=invis]
disturbance->middleman [constraint=false, arrowhead=diamond];

    middleman->biomass;
  
    evenness-> biomass;
    richness->biomass;

    nutrients -> middleman[arrowhead=none];
    nutrients -> evenness;
   nutrients -> richness;

 evenness-> richness [dir=both; minlen=3] ;

 { rank="same"; evenness, richness}
}
```

-   Investigate possible interaction between disturbance and nutrient addition on above ground biomass.
-   Use composite to model an interaction between nutrients and disturbance on biomass. 
-   Incorporate composite into full model
- How does the interaction term influence the R$^2$ of biomass?

## Exercise 5

::: {.column width="40%"}
Group = Disturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness;
    nutrients -> biomass;

    evenness-> biomass;
    richness->biomass;

 evenness-> richness [dir=both; minlen=3] ;

 { rank="same"; evenness, richness}

  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```
:::

::: {.column width="40%"}
Group = Undisturbed
```{dot}
//| fig-width: 5
digraph  {

   # a 'graph' statement
  graph [overlap = false, fontsize = 10, rankdir = TB, ranksep=.5]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica, color="gray"]
  
        evenness, richness;
        nutrients;

    nutrients->richness;
    nutrients -> evenness;
    nutrients -> biomass;

    evenness-> biomass;
    richness->biomass;

 evenness-> richness [dir=both; minlen=3] ;

 { rank="same"; evenness, richness}
  
 #{ rank="same"; precipitation, nutrients, disturbance}
}
```

:::


- Investigate possible interaction between disturbance and nutrient addition on biomass with a multi-group SEM.
- Test the hypothesis that all paths are the same for disturbed and undisturbed plots.
- Test the hypothesis that all paths are the same for disturbed and undisturbed plots.


<!-- ## Exercise 7

-   Account for nested experimental design with the `lavaan.survey` package:
    -   Add individual plots nested in fields.
 -->