---
title: "Case Study"
subtitle: "Sediment Respiration Pulses in Intermittent Rivers and Ephemeral Streams (Schiller et al., 2019)"
author: "Rebecca"
date: "10 November 2020"
output: html_document
---

#Intro
Most streams emit large quanitites of CO2 to the atmosphere, but so far contribution of intermittent rivers and ephemeral streams (IRES) were not included in these fluxes. 

#Methods
## Field and lab
A standardizes protocol was used to collect sediments and measure environmental variables from different regions of the world. Sediment samples were then analysed and measurements were taken upon rewetting.

## Statistical Analyses
In a first step, differences among measuring methods were tested. Second, the relationship between sediment respiration changes upon rewetting was measured using gls(). These first two steps are for now not of interest. 
The third step was to model causal relationships between environmental drivers and sediment respiration change upon rewetting using SEM, which is described in the following.

The authors distinguish between proximal drivers: Clay Content, Organic Carbon and C/N ratio, and distal drivers: Land use, Precipitation, Temperature, Riparian Cover and Channel Width.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# load packages
library(lavaan)
library (readxl)
library (spdep)
library(SoDA)
library(ade4)
library(adespatial)
library(ggplot2)
```


```{r}
# load data
setwd("C:/Users/Rebecca/polybox/PhD/UZH Courses/2020-SEM")
reaches <- read_excel("C:/Users/Rebecca/polybox/PhD/UZH Courses/2020-SEM/Data_Sediment Respiration.xlsx", 
                      sheet = "Data")

# create new variables Response.microresp, Response.bottle and Landuse

reaches$Response.microresp <- reaches$CO2MicrorespWet - reaches$CO2MicrorespDry # response of sediment respiration as difference between wet-dry for the Microresp method
reaches$Response.bottles <- reaches$CO2Bottles - reaches$CO2MicrorespDry # response of sediment respiration as difference between wet-dry for the Bottle method
reaches$Landuse <- reaches$PercentAgriculture + reaches$PercentUrban # Landuse as a measure of  anthropogenic area expressed as the sum of the percentages of agriculture area and urban area

```

```{r}
## create metamodel as described in the paper
metamodel <- '
Response.microresp ~ Organic Carbon + CarbonNitrogenRatio + ClayCoult
Organic Carbon ~ Landuse + AnnualPrecipitation + AnnualTemperature+RiparianCover
CarbonNitrogenRatio ~ Landuse+ Annual Precipitation + AnnualTemperature+ RiparianCover
RiparianCover ~ ActiveChannelWidth + Landuse
'
```

Exploratory data anlaysis suggested the need to square-root transform or ln-transform the response variables (including respiration rates) and five explanatory variables to ensure linearity of relationships and hence the suitability of the global estimation method. So let's see what we find:


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
## explanatory variables
hist(reaches$OrganicCarbon) ## left skewed
hist(reaches$CarbonNitrogenRatio) ## left skewed
hist(reaches$ClayCoult) ## left skewed
hist(reaches$RiparianCover) ## bimodal
hist(reaches$AnnualTemperature) ## normal
hist(reaches$AnnualPrecipitation) ## left skewed
hist(reaches$ActiveChannelWidth) ## left skewed
hist(reaches$Landuse) ## bimodal
## response variables
hist(reaches$Response.bottles) ## left skewed
hist(reaches$Response.microresp) ## left skewed
```


sem()

moindices()

