---
title: "Avanced structural equation modeling"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
header-includes:
   - \usepackage{svg}
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("here")
```



# To do
* non-linear relationships
* interactions
* multigroup fitting



# Load data and `lavaan`
Here, we use the very same data set as in the introductory part to SEM (`1_basic_modeling.html`).
Also, we rescale the biomass so that it is approximately in the same range as the other variables.


```{r}
library("lavaan")

seabloom <- read.table(here("2_Modeling/Data_preparation/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data.csv"),
                       sep = ",", header = TRUE)
# seabloom <- read.table("~/repos/Swiss_SEM/2_Modeling/Data_preparation/seabloom-2020-ele-dryad-data/cdr-e001-e002-output-data.csv",
#                        sep = ",", header = TRUE)
# seabloom <- seabloom[seabloom$year == 2000, ]

seabloom$mass.above <- seabloom$mass.above / 100
```




# Complex sampling structure
Often, we encounter more complex sampling or experimental schemes, e.g. the data is nested within sites or contains groups with non-random differences such as sexes or lifestages.
As a result, these data violate the principle of being [*i.i.d.* (independent and identically distributed)](https://en.wikipedia.org/wiki/Independent_and_identically_distributed_random_variables).
Thus, it is necessary to account for this structure in the data in the model.


The add-on package [`lavaan.survey`](https://cran.r-project.org/web/packages/lavaan.survey/index.html) allows the analysis of stratified, clustered or weighted data with the help of the package [`survey`](https://r-survey.r-forge.r-project.org/survey/).
With this, `lavaan` objects can be processed further with the specific data structure:
first, we initialize the design and then, post-process the `lavaan` object and compute the adjusted results.
The result is a corrected `lavaan` object.

As sampling was performed in three different fields (`A`, `B` and `C`), let's adjust for this structure.


## Rerun the model
First, we rerun the model. But this time, we will use the full data set over all the years. This increases sample size and allows to incorporate also precipitation.

![](Figures/SimplePrecSEM.svg){width=70%}


```{r}
simple <-
"mass.above ~ nadd + disk + precip.mm + rich + even
rich ~ nadd + precip.mm
even ~ nadd + precip.mm

rich ~~ even"

fit.simple <- sem(simple, data = seabloom, estimator = "MLM")
summary(fit.simple)

modindices(fit.simple, minimum.value = 3.84)
```


```{r}
fit.simple.up <- update(fit.simple, add = "rich ~ disk")
summary(fit.simple.up, fit.measures = TRUE)

modindices(fit.simple.up, minimum.value = 3)
```


```{r}
library("lavaan.survey")

design <- svydesign(ids = ~ field, nest = TRUE, data = seabloom)
fit.simple.up.nest <- lavaan.survey(lavaan.fit = fit.simple, 
                                    survey.design = design)
```

Two warnings pop up:

* The first one indicates that the `variance-covariance matrix of the estimated parameters (vcov) does not appear to be positive definite!`. However, the  smallest eigenvalue of -5.215416e-17 is very close to zero, what is likely due to  machine-precision.
* The second one might point to a sample size issue: for the year 2000 we have `r dim(seabloom)[1]` observations. Split into the three fields `A`, `B` and `C`, there are `r table(seabloom$field)[1:2]` and `r table(seabloom$field)[3]` observations in these three fields, respectively. With `r fit.simple.up.nest@Fit@npar` free parameters, this refers to `r round(table(seabloom$field)[1:2] / fit.simple.up.nest@Fit@npar, digits = 1)` and `r round(table(seabloom$field)[3] / fit.simple.up.nest@Fit@npar, digits = 1)`  samples per free parameter.
**This is actually low and might be the reason of the failed model fitting.**


```{r}
summary(fit.simple.up.nest, rsq = TRUE)
```



# Latent and composite variables
- difference between latent and composite variable
- SEM with latent variable(s)
- SEM with composite variable(s)

![](Figures/LatentCompositeSEM.svg){width=35%}


## Latent variable
Let's first create a latent variable "diversity" from the two measurements of diversity (richness and evenness).

```{r}
lv <- "diversity =~ rich + even"
cfa(lv, seabloom)
```


```{r}
lv <-"
# Latent variable definition
diversity =~ rich + even

mass.above ~ nadd + disk + diversity
diversity ~ nadd"

fit.lv <- sem(lv, data = seabloom, estimator = "MLM")
summary(fit.lv)
```


## Composite variable
Now, from the landuse parameters, we can create a composite variable...

```{r}
comp <- '
landuse <~ disk + nadd

rich ~ nadd
even ~ nadd

mass.above ~ rich + even + landuse'


fit.comp <- sem(comp, data = seabloom, std.lv = TRUE)

```



<!-- ```{r} -->
<!-- compositeModel <- ' -->
<!--     #1) define the composite, scale to logN -->
<!--     Nitrogen ~ 1*logN + logNcen2 #loading on the significant path! -->
<!--   -->
<!--     #2) Specify 0 error variance -->
<!--     Nitrogen ~~ 0*Nitrogen -->
<!--  -->
<!--       #3) now, because we need to represent this as a latent variable -->
<!--       #show how species richness is an _indicator_ of nitrogen -->
<!--       Nitrogen =~ SA -->
<!--   -->
<!--     #4) BUT, make sure the variance of SA is estimated -->
<!--       SA ~~ SA -->
<!--   -->
<!--    #Regional Richness also has an effect -->
<!--     SA ~ SR -->
<!--   -->
<!--     #And account for the derivation of the square term from the linear term -->
<!--     logNcen2 ~ logN -->
<!--       ' -->
<!--   -->
<!--  # we specify std.lv=T so that the Nitrogen-SA relationship isn't fixed to 1 -->
<!--  compositeFit <- sem(compositeModel, data = seabloom, std.lv = TRUE) -->
<!--  -->
<!-- ```  -->


## Model pruning
Now, let's delete all non-significant paths from the model to obtain the most parsimonous model.
This, however, leads to a poor model fit.


```{r}
sem.prune <-
"mass.above ~ nadd + rich
rich ~ nadd
even ~ nadd 

rich ~~ even"

fit.sem.prune <- sem(sem.prune, data = seabloom)
summary(fit.sem.prune, rsq = TRUE)
```


# Interactions
![](Figures/InteractionSEM.svg){width=70%}


# Autocorrelation (temporal, spatial)


# Test of mediation


