---
title: "Plotting directed acyclic graphs with the DiagrammR pacge"
author: "Frank Pennekamp"
date: "11/12/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Drawing simple DAGs with DiagrammR

* This document contains code and illustrations how to draw directed acyclic graphs with the DiagrammR package in R
* The DiagrammR package uses the dot language of the GraphViz package
* GraphViz pocket reference: https://graphs.grevian.org/
* Official GraphViz page: https://graphviz.org/about/

## Linear model

```{r lm}
library(DiagrammeR)

grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X; Y;

  # 'edge' statements
  X -> Y
}
")
```


## Multiple regression

```{r lm2}
grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X_1; X_2; Y

  # several 'edge' statements
  X_1 -> Y
  X_2 -> Y
}
")
```

## Regression with interaction efefct

```{r interact}
grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X_1; X_2; Y, middleman

  # several 'edge' statements
    X_1->middleman[arrowhead=none];
    middleman->Y;
    middleman[shape=none width=0 height=0 label='' style=invis]
    X_2->middleman [constraint=false];

}
")
```

## More than one response (indirect effects)

```{r indirect}
grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X; Y, Z

  # several 'edge' statements
    X->Y[label='0.2'];
    Y->Z[label='0.2'];
}
")
```

## More than one response (direct and indirect effects)

```{r direct}
grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10, layout = dot,
       rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  X; Y, Z

  # several 'edge' statements
    X->Y;
    Y->Z;
    X->Z;
}
")
```

## Latent variable (circle)

```{r latent}
grViz("
digraph boxes_and_circles {
   
   node [shape=box]
   X1;X2;X3;Z;

   
   node [shape=circle]
   Y
    
    Y -> X1;
    Y -> X2;
    Y -> X3;
    Y -> Z;
    Z;
    
  X1->X2->X3 [style='invis']
  {rank = 'max'; Z}
  {rank = 'min'; X1}
  {rank = 'same'; X1;X2;X3}
}
")
```


## Composite variable (hexagon)

```{r composite}
grViz("
digraph boxes_and_circles {
   
   node [shape=box]
   X1;X2;X3;Z;

   
   node [shape=hexagon]
   Y
    
      X1 -> Y;
    X2 -> Y;
    X3 -> Y;
    Y -> Z;
    Z;
    
  X1->X2->X3 [style='invis']
  {rank = 'max'; Z}
  {rank = 'same'; X1;X2;X3}
}
")
```


## Put it all together and make it prettier


### Example 1
```{r labelled}
grViz('
digraph boxes_and_circles {
   
      graph[layout=dot]
      
    subgraph {
    
    node[shape=box]
    a -> b[color=black, penwidth=3.0, label="0.5"]
    a -> c[color=black, penwidth=3.0, label="0.5"]
    c -> b[color=red, penwidth=1.5, label="-0.25"]
    c -> e[color=red, penwidth=1.5, label="-0.25"]
      }
      
      
    subgraph {
           
    node[shape=box]

    e -> e[color=black, penwidth=3.0, label="0.5"]
    e -> b[color=red, penwidth=1.5, label="-0.25"]

      }
       
      
{rank = "same"; c;e}


}
')
```

### Example 2

```{r}
grViz("
digraph {

  graph [ranksep = 1]
  'i' [shape = 'circle']
  's' [shape = 'circle']
  't1' [shape = 'square']
  't2' [shape = 'square']
  't3' [shape = 'square']
  't4' [shape = 'square']
  'x1' [shape = 'square']
  'x2' [shape = 'square']
  'c1' [shape = 'square']
  'c2' [shape = 'square']
  'c3' [shape = 'square']
  'c4' [shape = 'square']
  'i'->'t1' [style = 'dashed', label = '1']
  'i'->'t2' [style = 'dashed', label = '1']
  'i'->'t3' [style = 'dashed', label = '1']
  'i'->'t4' [style = 'dashed', label = '1']
  's'->'t1' [style = 'dashed', label = '0']
  's'->'t2' [style = 'dashed', label = '1']
  's'->'t3' [style = 'dashed', label = '2']
  's'->'t4' [style = 'dashed', label = '3']
  'x1'->'i' [style = 'solid', label = '0.61']
  'x2'->'i' [style = 'solid', label = '0.6']
  'x1'->'s' [style = 'solid', label = '0.26']
  'x2'->'s' [style = 'solid', label = '0.52']
  'c1'->'t1' [style = 'solid', label = '0.14']
  'c2'->'t2' [style = 'solid', label = '0.29']
  'c3'->'t3' [style = 'solid', label = '0.33']
  'c4'->'t4' [style = 'solid', label = '0.33']

  # Additional constraints on the graph
  t1->t2->t3->t4 [style='invis']
  {rank = 'max'; c1; c2; c3; c4;}
  {rank = 'same'; t1; t2; t3; t4;}
}
")
```


